Tidy Data

### 2013-06-05 tidy-data.tex
Changes for JSS style

%% almost as usual
\author{Hadley Wickham\\RStudio}
\title{Tidy data}

\Plainauthor{Hadley Wickham}
\Plaintitle{Tidy data}

@@

\DeclareGraphicsExtensions{.png,.pdf,.jpg}
\usepackage[format=plain,font=small]{caption}
\usepackage{booktabs}
\usepackage{verbatim}
\usepackage{epigraph}
\usepackage[subrefformat=parens]{subfig}

@@

The development of tidy data has been driven by my struggles working with real-world datasets, which are often organised in bizarre ways. I have spent countless hours struggling to get these datasets organised in a way that makes data analysis possible, let alone easy. I have also struggled to impart these skills to my students, so they can tackle problems in the real-world, where there are few constraints on how datasets are organised. In the course of these struggles I developed the reshape and reshape2 \citep{wickham:2007b} R packages, but while I could intuitively use the tools, I lacked the framework to make my understanding explicit. This paper provides that framework, and provides a consistent ``philosophy of data'' that underlies my work in the plyr \citep{me:plyr} and ggplot2 \citep{me:ggplot2} packages.

@@

Statistical datasets usually come in the form of a rectangular \textbf{table}, made up of \textbf{rows} and \textbf{columns}. Tables are usually labelled with column names, and sometimes with row names. The table is a collection of data \textbf{value}s, typically either numeric (if quantitative) or character (if qualitative). Values are grouped into \textbf{variables}, measurements of a single quantifiable property. Multiple measurements made on the same \textbf{observational unit} form an \textbf{observation}.

@@

It's not always obvious what qualifies as a variable, and a variable in one analysis may be a value in another. For example, if the columns in the example were were height and weight we would have been happy to call them variables. If the columns were height and width, it would be less clear cut, as we might think of height and weight as values of a dimension variable. If the columns were home phone and work phone, we could treat these as two variables, but in a fraud detection environment we might want variables phone number and number type because the use of one phone number for multiple people might suggest fraud. A general rule of thumb is that it is easier to describe functional relationships between variables/columns (e.g.\ z is a linear combination of x and y, density is the ratio of weight to volume) than between rows, and it is easier to make comparisons between groups of rows/observations (e.g.\ average of group a vs.\ average of group b) than between columns. If you find yourself fighting these constraints, you may not have correctly identified the appropriate variables for your problem.

In a given analysis, there may also be multiple \textbf{types} of observational unit. For example, in a trial of new allergy medication we might have three observational types: demographic data collected on \emph{each person} (age, sex, race), on \emph{each person on each day} of hay fever season (number of sneezes, redness of eyes), and on \emph{each day} (temperature, pollen count).

A dataset is messy or tidy depending on how rows, columns and tables are matched up with observations, variables and types. In {\bf tidy data}:

@@

\noindent This is Codd's 3rd normal form \citep{codd:1990}, but with the constraints framed in statistical language, and the focus on a single dataset, not the many connected datasets common in relational databases. I will call other arrangements of data {\bf messy}.

@@

While order of variables and observations does not affect analysis, a good ordering makes it easier to scan the raw values. One way of organising variables is by their role in the analysis: are values fixed by the design of the data collection, or are they measured during the course of the experiment? \textbf{Fixed} variables describe the experimental design and are known in advance. These are often called dimensions by computer scientists, and typically denoted by subscripts on random variables. \textbf{Measured} variables are what we actually measure in the study. Fixed variables should come first, followed by measured variables, each ordered so that related variables are contiguous. Rows can then be ordered by the first variable, breaking ties with the second and subsequent variables. This is the convention adopted by all tabular displays in this paper. When using tables for communication, this so-called ``Alabama first'' \citep{wainer:2000} ordering should be abandoned, and replaced with an ordering based on a meaningful variable.

@@

Surprisingly, most messy datasets, including types of messiness not explicitly described above, can be tidied with a small set of tools: ``melting'', string splitting, and ``casting''. The following sections illustrate each problem with a real dataset that I have encountered in practice, and show how to tidy it. The complete datasets and the R code used to tidy them are available online at \url{https://github.com/hadley/tidy-data}, and in the online supplementary materials for this paper.

@@

The Pew dataset has one colvar, religion, and melting yields Table~\ref{tbl:pew-clean}. The column variable has been renamed to {\tt income}, and the value column to {\tt freq}, to better reflect their roles in this dataset. This form is tidy: each column represents a variable and each row represents an observation, in this case a demographic unit corresponding to a combination of religion and income.

@@

Another common use of this data format is to record regularly spaced observations over time. For example, the billboard dataset shown in Table~\ref{tbl:billboard-raw} records the date a song first entered the billboard top 100. It has variables artist, track, date.entered, rank and week. The rank in each week after it enters the top 100 is recorded in 75 columns, {\tt wk1} to {\tt wk75}. If a song is in the top 100 for less than 75 weeks the remaining columns are filled with missing values.  This form of storage is not tidy, but is useful for data entry because it reduces duplication: otherwise each song in each week would need its own row, and song metadata like title and artist would need to be repeated. This issue will be discussed in more depth in Section~\ref{sub:multiple-types}.

\begin{table}[htbp]
  \centering
  \input{data/billboard-raw.tex}
  \caption{The first eight billboard top hits for 2000. Other columns not shown are {\tt wk4}, {\tt wk5}, ..., {\tt wk75}.}
  \label{tbl:billboard-raw}
\end{table}

@@

\begin{table}[htbp]
  \centering
  \input{data/billboard-clean.tex}
  \caption{First fifteen rows of the tidied billboard dataset. The {\tt date} column does not appear in the original table, but can be computed from {\tt date.entered} and {\tt week}.}
  \label{tbl:billboard-clean}
\end{table}

@@

\begin{table}[htbp]
  \centering
  \input{data/tb-raw.tex}
  \caption{Original TB dataset. Corresponding to each `m' column for males, there is also an `f' column for females: {\tt f1524}, {\tt f2534} and so on. These are not shown to conserve space. Note the mixture of 0s and missing values (---): this is due to the data collection process and the distinction is important for this dataset.}
  \label{tbl:tb-raw}
\end{table}

Column headers in this format are often separated by some character ({\tt .}, {\tt -}, {\tt \_}, {\tt :}) and the string can be broken into pieces using that character as a divider. In other cases, such as for this dataset, more careful string processing is required, or the variable names need to be matched to a lookup table that converts single compound values to values of multiple variables.

@@

This dataset is mostly tidy, but we have two variables stored in rows: {\tt tmin} and {\tt tmax}, the type of observation. Not shown in this example are other meteorological variables prcp (precipitation) and snow (snowfall). Fixing this requires the cast, or unstack, operation, which performs the inverse of melting, rotating the {\tt element} variable back out into the columns to give Table~\subref*{tbl:weather-tidy}. This form is tidy: there is one variable in each column, and each row represents a day's observations. The cast operation is described in depth in \citet{wickham:2007b}.

\begin{table}[htbp]
  \centering
  \input{data/weather-raw.tex}
  \caption{Original weather dataset.  There is a column for each possible day in the month.  Columns {\tt d11} to {\tt d31} have been omitted to conserve space.}
  \label{tbl:weather-raw}
\end{table}

@@

  \caption{(a) Molten weather dataset.  This is almost tidy, but the {\tt element} column contains names of variables, not values. Missing values are dropped to conserve space. (b) Tidy weather dataset. Each row representation the meteorological measurements for a single day.  There are two measured variables, minimum ({\tt tmin}) and maximum ({\tt tmax}) temperature; all other variables are fixed.}
  \label{tbl:weather-clean}
\end{table}

@@

  \caption{Normalised billboard dataset split up into song dataset (left) and rank dataset (right). First 15 rows of each dataset shown; {\tt genre} omitted from song dataset, {\tt week} omitted from rank dataset.}
  \label{tbl:billboard-normal}
\end{table}

@@

Next, I give examples of tidy and messy tools for three important components of analysis: data manipulation, visualisation and modelling. I will focus particularly on tools provided by R \citep{R}, because it has many existing tidy tools, but I will also touch on other statistical programming environments.

@@

Data manipulation includes variable-by-variable transformation (e.g. {\tt log} or {\tt sqrt}), as well as aggregation, filtering and reordering. In my experience, there are four extremely common operations that are performed over and over again in the course of an analysis. These are the four fundamental verbs of data manipulation:

@@

  \item Transforming, where new variables are added or existing variables
  modified. These modifications can involve either a single variable (e.g.\ 
  log-transforming a variable), or involve multiple variables (e.g.\ computing
  density from weight and volume).

  \item Aggregating, where multiple values are collapsed into a single value,
  e.g.\ by summing or taking means.

@@

In R, filtering and transforming are performed by the base R functions {\tt subset} and {\tt transform}. These are input and output-tidy. The {\tt aggregate} function performs group-wise aggregation, is input-tidy, and is output-tidy providing a single aggregation method is used. The plyr package provides a tidy {\tt summarise} and {\tt arrange} function for aggregation and sorting.

The four verbs can be, and often are, modified by the \textbf{by} preposition. We often need group-wise aggregates, transformations or subsets: pick the biggest in each group, average over replicates and so on. Combining each of the four verbs with a by operator allows them to operate on subsets of a data frame at a time. Many {\sc sas} {\sc proc}s possess a {\sc by} statement which allows the operation to be performed by group, and are generally input-tidy. Base R possesses a {\tt by} function, which is input-tidy, but not output-tidy, because it produces a list. The {\tt ddply} function from the {\tt plyr} package is a tidy alternative.

@@

Other tools are needed when we have multiple datasets. An advantage of tidy data is the ease with which it can be combined with other tidy datasets: we just need a {\bf join} operator which works by matching common variables and adding new columns. This is implemented in the {\tt merge} function in base R, or the {\tt join} function in {\tt plyr}. Compare this to the difficulty of combining datasets stored in arrays; these typically require painstaking alignment before matrix operations can be used, and errors can be very hard to detect.

@@

Tidy visualisation tools need only to be input-tidy as their output is visual. Domain specific languages work particularly well for the visualisation of tidy datasets because they can describe a visualisation as a mapping between variables in the dataset and aesthetic properties of the graphic like position, size, shape and colour. This is the idea behind the grammar of graphics \citep{wilkinson:2006}, and the layered grammar of graphics \citep{wickham:2007d}, an extension tailored specifically for R.

Most graphical tools in R are input tidy, including the base {\tt plot} function, the lattice family of plots \citep{sarkar:2008} and ggplot2 \citep{me:ggplot2}. Some specialised tools exist for visualising messy datasets. Some base R functions like {\tt barplot}, {\tt matplot}, {\tt dotchart}, and {\tt mosaicplot}, work with messy datasets where variables are spread out over multiple columns. Similarly, the parallel coordinates plot \citep{wegman:1990,inselberg:1985} can be used to create time series plots for messy datasets where each time point is a column.

@@

  \item R (lm): \verb|y ~ a + b + c * d|

  \item SAS (PROC GLM): \verb|y = a + b + c * d|

  \item SPSS (glm): \verb|y BY a b c d / DESIGN a b c d c*d |

@@

There have been some attempts to adapt modelling functions for specific types of messy datasets. For example, in {\sc sas}'s {\sc proc glm}, multiple variables on the response side of the equation will be interpreted as repeated measures if the {\sc repeated} keyword is present. For the raw billboard data, we could construct a model of the form \verb|wk1-wk76 = track| instead of \verb|rank = week * track| on the tidy data (provided week is labelled as a categorical variable).

Another interesting example is the classic paired t-test, which can be computed in two ways depending on how the data is stored. If the data is stored as in Table~\subref*{tbl:paired}, then a paired t-test is just a t-test applied to the mean difference between x and y. If it is stored in the form of Table~\subref*{tbl:mixed}, then an equivalent model is to fit a mixed effects model, with a fixed dummy variable representing the ``variable'', and a random intercept for each id. (In R's lmer4 notation, this is \verb#value ~ variable + (1 | id)#). Either way of modelling the data yields the same result. Without more information we can't say which form of the data is tidy: if x and y represent length of left and right arms, then Table~\subref*{tbl:paired} is tidy, if x and y represent measurements on day 1 and day 10, then Table~\subref*{tbl:mixed} is tidy.

@@

While model inputs usually require tidy inputs, model outputs, such as predictions and estimated coefficients, aren't always tidy. This makes it more difficult to combine results from multiple models. For example, in R, the default representation of model coefficients is not tidy because it does not have an explicit variable that records the variable name for each estimate: these are instead recorded as row names. In R, row names must be unique, so combining coefficients from many models (e.g.\ from bootstrap resamples, or subgroups) requires workarounds to avoid losing important information.  This knocks you out of the flow of analysis and makes it harder to combine the results from multiple models.

@@

The following case study illustrates how tidy data and tidy tools make data analysis easier by easing the transitions between manipulation, visualisation and modelling: you will not see any code that exists solely to get the output of one function into the right format for input to another. I'll show the R code that performs the analysis, but even if you're not familiar with R or the exact idioms I use, I've tried to make it easy to understand by tightly interleaving code, results and explanation.

@@

To achieve our goal of finding unusual time courses, we'll start by counting the number of deaths in each hour ({\tt hod}) for each cause ({\tt cod}) with the tidy {\tt count} function. Then we remove missing (and hence uninformative for our purpose) values with {\tt subset}. This gives Table~\subref*{tbl:counts:1}.

\begin{verbatim}
# Count the the number of deaths in each hour of the day for each 
# cause of death.  
hod2 <- count(deaths, c("hod", "cod"))

# Remove missing observations
hod2 <- subset(hod2, !is.na(hod))
\end{verbatim}

Next we join the dataset to {\tt codes} dataset that gives informative labels for the disease. These two datasets are connected by the {\tt cod} variable. This adds a new variable, {\tt disease}, shown in Table~\subref*{tbl:counts:2}).

\begin{verbatim}
hod2 <- join(hod2, codes, by = "cod")
\end{verbatim}

The total deaths for each cause vary over several orders of magnitude: there were 46,794 deaths from heart attack, and 10 deaths from avalanche. This means that it makes more sense to compare the proportion of deaths in each hour, rather than the total number. We compute this by breaking the dataset down by {\tt cod}, and then {\tt transform}ing to add a new {\tt prop} column, the hourly frequency divided by the total number of deaths from that cause. This new columns is Table~\subref*{tbl:counts:3}.

\begin{verbatim}
# ddply breaks down its first argument (hod2) by its second argument (the
# cod variable), and apply its third argument (transform) to each piece. The
# fourth argument (prop = freq / sum(freq)) is passed on to transform.
hod2 <- ddply(hod2, "cod", transform, prop = freq / sum(freq))
\end{verbatim}

We then compute the overall average death rate for each hour, and merge that back into the original dataset. This yields Table~\subref*{tbl:counts:4} and allows us to easily compare each disease with the overall pattern by comparing {\tt prop} to {\tt prop\_all}.

\begin{verbatim}
# Work out number of people dying each hour by breaking down hod2
# by hod, and summarising by computing total over each cause of death
overall <- ddply(hod2, "hod", summarise, freq_all = sum(freq))

# Work out overall proportion of people dying in each hour
overall <- transform(overall, prop_all = freq_all / sum(freq_all))

# Join overall dataset with individual dataset to make it easier to
# compare the two.
hod2 <- join(hod2, overall, by = "hod")
\end{verbatim}

@@

  \caption{A sample of four diseases and four hours from {\tt hod2} data frame.}
  \label{tbl:counts}
\end{table}

Next we compute a distance between the time course of each cause of death and the overall time course. Here we use a simple mean squared deviation.  We also record the sample size, the total number of deaths from that cause.

\begin{verbatim}
# Compute deviation for each disease by summarising with the number
# of deaths, and mean squared deviation from the overall trend.
devi <- ddply(hod2, "cod", summarise, 
  n = sum(freq), 
  dist = mean((prop - prop_all)^2))

# Only work with diseases with more than 50 deaths (~2/hour)
devi <- subset(devi, n > 50)
\end{verbatim}

We don't know the variance characteristics of this estimator, so we explore it visually by plotting n vs.\ deviation, Figure~\ref{fig:deviation-raw}. The linear scale shows little, except the variability decreases with sample size, but on the log-log scale, Figure~\ref{fig:deviation-log} there is a clear pattern.

\begin{verbatim}
# Visualise the devi dataset,
ggplot(data = devi, aes(x = n, y = dist) + geom_point()

# Modify the previous plot, converting to log scales, and adding a
# smoother from a robust linear model.
last_plot() + 
  scale_x_log10() + 
  scale_y_log10() +
  geom_smooth(method = "rlm", se = F)
\end{verbatim}

@@

We are interested in points that are high relative to other causes with the same number of deaths: these are diseases most different from the overall pattern. To find these unusual points, we fit a robust linear model and plot the residuals, as shown in Figure~\ref{fig:devi-resid}. The plot shows an empty region around a residual of 1.5, so somewhat arbitrarily we select those diseases with a residual greater than 1.5. We do that in two steps: first selecting the appropriate rows from {\tt devi} (one row per disease), and then finding the matching time course information in the original summary dataset (24 rows per disease).

\begin{verbatim}
# Compute residuals from robust linear model on log transformed values
devi$resid <- resid(rlm(log(dist) ~ log(n), data = devi))

# Select diseases with residuals bigger than 1.5
unusual <- subset(devi, resid > 1.5)

# Extract matching time course dataset for unusual diseases
hod_unusual <- match_df(hod2, unusual)
\end{verbatim}

Finally, we plot the time courses of the unusual causes in Figure~\ref{fig:disease}. We break the diseases into two plots because of the differences in variability: the top plot shows diseases with over 350 deaths and the bottom with less than 350. The causes of death fall in to three main groups: murders, drowning, and transportation related. Murders are more common at night, drowning in the afternoon, and transportation related during commute times.

\begin{verbatim}
# Visualise unusual deaths dataset hod on x-axis, prop on y-axis, display by line, with pale gray
# line showing overall death rate.
ggplot(data = subset(hod_unusual, n > 350), aes(x = hod, y = prop)) + 
 geom_line(aes(y = prop_all), data = overall, colour = "grey50") +
 geom_line() + 
 facet_wrap(~ disease, ncol = 3)
\end{verbatim}

>>>>

%% almost as usual
\author{Hadley Wickham\\RStudio}
\title{Tidy Data}

\Plainauthor{Hadley Wickham}
\Plaintitle{Tidy Data}

@@

\DeclareGraphicsExtensions{.png,.pdf,.jpg}
\usepackage[format=plain,font=small]{caption}
\usepackage{booktabs}
\usepackage{fancyvrb}
\renewcommand{\FancyVerbFormatLine}[1]{R> #1}
\usepackage{epigraph}
\usepackage[subrefformat=parens]{subfig}

@@

The development of tidy data has been driven by my struggles working with real-world datasets, which are often organised in bizarre ways. I have spent countless hours struggling to get these datasets organised in a way that makes data analysis possible, let alone easy. I have also struggled to impart these skills to my students, so they can tackle problems in the real-world, where there are few constraints on how datasets are organised. In the course of these struggles I developed the \pkg{reshape} and \pkg{reshape2} \citep{wickham:2007b} \proglang{R} packages, but while I could intuitively use the tools, I lacked the framework to make my understanding explicit. This paper provides that framework, and provides a consistent ``philosophy of data'' that underlies my work in the \pkg{plyr} \citep{me:plyr} and \pkg{ggplot2} \citep{me:ggplot2} packages.

@@

Statistical datasets usually come in the form of a rectangular table, made up of rows and columns. Tables are usually labelled with column names, and sometimes with row names. The table is a collection of data values, typically either numeric (if quantitative) or character (if qualitative). Values are grouped into variables, measurements of a single quantifiable property. Multiple measurements made on the same observational unit form an observation.

@@

It's not always obvious what qualifies as a variable, and a variable in one analysis may be a value in another. For example, if the columns in the example were were height and weight we would have been happy to call them variables. If the columns were height and width, it would be less clear cut, as we might think of height and weight as values of a dimension variable. If the columns were home phone and work phone, we could treat these as two variables, but in a fraud detection environment we might want variables phone number and number type because the use of one phone number for multiple people might suggest fraud. A general rule of thumb is that it is easier to describe functional relationships between variables/columns (e.g., z is a linear combination of x and y, density is the ratio of weight to volume) than between rows, and it is easier to make comparisons between groups of rows/observations (e.g., average of group a vs.\ average of group b) than between columns. If you find yourself fighting these constraints, you may not have correctly identified the appropriate variables for your problem.

In a given analysis, there may also be multiple types of observational unit. For example, in a trial of new allergy medication we might have three observational types: demographic data collected on \emph{each person} (age, sex, race), on \emph{each person on each day} of hay fever season (number of sneezes, redness of eyes), and on \emph{each day} (temperature, pollen count).

A dataset is messy or tidy depending on how rows, columns and tables are matched up with observations, variables and types. In tidy data:

@@

\noindent This is Codd's 3rd normal form \citep{codd:1990}, but with the constraints framed in statistical language, and the focus on a single dataset, not the many connected datasets common in relational databases. I will call other arrangements of data messy.

@@

While order of variables and observations does not affect analysis, a good ordering makes it easier to scan the raw values. One way of organising variables is by their role in the analysis: are values fixed by the design of the data collection, or are they measured during the course of the experiment? Fixed variables describe the experimental design and are known in advance. These are often called dimensions by computer scientists, and typically denoted by subscripts on random variables. Measured variables are what we actually measure in the study. Fixed variables should come first, followed by measured variables, each ordered so that related variables are contiguous. Rows can then be ordered by the first variable, breaking ties with the second and subsequent variables. This is the convention adopted by all tabular displays in this paper. When using tables for communication, this so-called ``Alabama first'' \citep{wainer:2000} ordering should be abandoned, and replaced with an ordering based on a meaningful variable.

@@

Surprisingly, most messy datasets, including types of messiness not explicitly described above, can be tidied with a small set of tools: ``melting'', string splitting, and ``casting''. The following sections illustrate each problem with a real dataset that I have encountered in practice, and show how to tidy it. The complete datasets and the \proglang{R} code used to tidy them are available online at \url{https://github.com/hadley/tidy-data}, and in the online supplementary materials for this paper.

@@

The Pew dataset has one colvar, religion, and melting yields Table~\ref{tbl:pew-clean}. The column variable has been renamed to \code{income}, and the value column to \code{freq}, to better reflect their roles in this dataset. This form is tidy: each column represents a variable and each row represents an observation, in this case a demographic unit corresponding to a combination of religion and income.

@@

Another common use of this data format is to record regularly spaced observations over time. For example, the billboard dataset shown in Table~\ref{tbl:billboard-raw} records the date a song first entered the billboard top 100. It has variables artist, track, date.entered, rank and week. The rank in each week after it enters the top 100 is recorded in 75 columns, \code{wk1} to \code{wk75}. If a song is in the top 100 for less than 75 weeks the remaining columns are filled with missing values.  This form of storage is not tidy, but is useful for data entry because it reduces duplication: otherwise each song in each week would need its own row, and song metadata like title and artist would need to be repeated. This issue will be discussed in more depth in Section~\ref{sub:multiple-types}.

\begin{table}[htbp]
  \centering
  \input{data/billboard-raw.tex}
  \caption{The first eight billboard top hits for 2000. Other columns not shown are \code{wk4}, \code{wk5}, ..., \code{wk75}.}
  \label{tbl:billboard-raw}
\end{table}

@@

\begin{table}[htbp]
  \centering
  \input{data/billboard-clean.tex}
  \caption{First fifteen rows of the tidied billboard dataset. The \code{date} column does not appear in the original table, but can be computed from \code{date.entered} and \code{week}.}
  \label{tbl:billboard-clean}
\end{table}

@@

\begin{table}[htbp]
  \centering
  \input{data/tb-raw.tex}
  \caption{Original TB dataset. Corresponding to each `m' column for males, there is also an `f' column for females: \code{f1524}, \code{f2534} and so on. These are not shown to conserve space. Note the mixture of 0s and missing values (---): this is due to the data collection process and the distinction is important for this dataset.}
  \label{tbl:tb-raw}
\end{table}

Column headers in this format are often separated by some character (\code{.}, \code{-}, \code{\_}, \code{:}) and the string can be broken into pieces using that character as a divider. In other cases, such as for this dataset, more careful string processing is required, or the variable names need to be matched to a lookup table that converts single compound values to values of multiple variables.

@@

This dataset is mostly tidy, but we have two variables stored in rows: \code{tmin} and \code{tmax}, the type of observation. Not shown in this example are other meteorological variables prcp (precipitation) and snow (snowfall). Fixing this requires the cast, or unstack, operation, which performs the inverse of melting, rotating the \code{element} variable back out into the columns to give Table~\subref*{tbl:weather-tidy}. This form is tidy: there is one variable in each column, and each row represents a day's observations. The cast operation is described in depth in \citet{wickham:2007b}.

\begin{table}[htbp]
  \centering
  \input{data/weather-raw.tex}
  \caption{Original weather dataset.  There is a column for each possible day in the month.  Columns \code{d11} to \code{d31} have been omitted to conserve space.}
  \label{tbl:weather-raw}
\end{table}

@@

  \caption{(a) Molten weather dataset.  This is almost tidy, but the \code{element} column contains names of variables, not values. Missing values are dropped to conserve space. (b) Tidy weather dataset. Each row representation the meteorological measurements for a single day.  There are two measured variables, minimum (\code{tmin}) and maximum (\code{tmax}) temperature; all other variables are fixed.}
  \label{tbl:weather-clean}
\end{table}

@@

  \caption{Normalised billboard dataset split up into song dataset (left) and rank dataset (right). First 15 rows of each dataset shown; \code{genre} omitted from song dataset, \code{week} omitted from rank dataset.}
  \label{tbl:billboard-normal}
\end{table}

@@

Next, I give examples of tidy and messy tools for three important components of analysis: data manipulation, visualisation and modelling. I will focus particularly on tools provided by \proglang{R} \citep{R}, because it has many existing tidy tools, but I will also touch on other statistical programming environments.

@@

Data manipulation includes variable-by-variable transformation (e.g., \code{log} or \code{sqrt}), as well as aggregation, filtering and reordering. In my experience, there are four extremely common operations that are performed over and over again in the course of an analysis. These are the four fundamental verbs of data manipulation:

@@

  \item Transforming, where new variables are added or existing variables
  modified. These modifications can involve either a single variable (e.g., 
  log-transforming a variable), or involve multiple variables (e.g., computing
  density from weight and volume).

  \item Aggregating, where multiple values are collapsed into a single value,
  e.g., by summing or taking means.

@@

In \proglang{R}, filtering and transforming are performed by the base \proglang{R} functions \code{subset} and \code{transform}. These are input and output-tidy. The \code{aggregate} function performs group-wise aggregation, is input-tidy, and is output-tidy providing a single aggregation method is used. The \pkg{plyr} package provides a tidy \code{summarise} and \code{arrange} function for aggregation and sorting.

The four verbs can be, and often are, modified by the by preposition. We often need group-wise aggregates, transformations or subsets: pick the biggest in each group, average over replicates and so on. Combining each of the four verbs with a by operator allows them to operate on subsets of a data frame at a time. Many \proglang{SAS} {\sc proc}s possess a {\sc by} statement which allows the operation to be performed by group, and are generally input-tidy. Base \proglang{R} possesses a \code{by} function, which is input-tidy, but not output-tidy, because it produces a list. The \code{ddply} function from the \pkg{plyr} package is a tidy alternative.

@@

Other tools are needed when we have multiple datasets. An advantage of tidy data is the ease with which it can be combined with other tidy datasets: we just need a join operator which works by matching common variables and adding new columns. This is implemented in the \code{merge} function in base \proglang{R}, or the \code{join} function in \pkg{plyr}. Compare this to the difficulty of combining datasets stored in arrays; these typically require painstaking alignment before matrix operations can be used, and errors can be very hard to detect.

@@

Tidy visualisation tools need only to be input-tidy as their output is visual. Domain specific languages work particularly well for the visualisation of tidy datasets because they can describe a visualisation as a mapping between variables in the dataset and aesthetic properties of the graphic like position, size, shape and colour. This is the idea behind the grammar of graphics \citep{wilkinson:2006}, and the layered grammar of graphics \citep{wickham:2007d}, an extension tailored specifically for \proglang{R}.

Most graphical tools in \proglang{R} are input tidy, including the \pkg{base} \code{plot} function, the \pkg{lattice} family of plots \citep{sarkar:2008} and \pkg{ggplot2} \citep{me:ggplot2}. Some specialised tools exist for visualising messy datasets. Some base \proglang{R} functions like \code{barplot}, \code{matplot}, \code{dotchart}, and \code{mosaicplot}, work with messy datasets where variables are spread out over multiple columns. Similarly, the parallel coordinates plot \citep{wegman:1990,inselberg:1985} can be used to create time series plots for messy datasets where each time point is a column.

@@

  \item \proglang{R} (\code{lm}): \code{y ~ a + b + c * d}

  \item \proglang{SAS} (\code{PROC GLM}): \code{y = a + b + c * d}

  \item \proglang{SPSS} (\code{glm}): \code{y BY a b c d / DESIGN a b c d c*d}

@@

There have been some attempts to adapt modelling functions for specific types of messy datasets. For example, in \proglang{SAS}'s \code{proc glm}, multiple variables on the response side of the equation will be interpreted as repeated measures if the {\sc repeated} keyword is present. For the raw billboard data, we could construct a model of the form \code{wk1-wk76 = track} instead of \code{rank = week * track} on the tidy data (provided week is labelled as a categorical variable).

Another interesting example is the classic paired t-test, which can be computed in two ways depending on how the data is stored. If the data is stored as in Table~\subref*{tbl:paired}, then a paired t-test is just a t-test applied to the mean difference between x and y. If it is stored in the form of Table~\subref*{tbl:mixed}, then an equivalent model is to fit a mixed effects model, with a fixed dummy variable representing the ``variable'', and a random intercept for each id. (In \proglang{R}'s lmer4 notation, this is \code{value ~ variable + (1 | id)}). Either way of modelling the data yields the same result. Without more information we can't say which form of the data is tidy: if x and y represent length of left and right arms, then Table~\subref*{tbl:paired} is tidy, if x and y represent measurements on day 1 and day 10, then Table~\subref*{tbl:mixed} is tidy.

@@

While model inputs usually require tidy inputs, model outputs, such as predictions and estimated coefficients, aren't always tidy. This makes it more difficult to combine results from multiple models. For example, in \proglang{R}, the default representation of model coefficients is not tidy because it does not have an explicit variable that records the variable name for each estimate: these are instead recorded as row names. In \proglang{R}, row names must be unique, so combining coefficients from many models (e.g., from bootstrap resamples, or subgroups) requires workarounds to avoid losing important information.  This knocks you out of the flow of analysis and makes it harder to combine the results from multiple models.

@@

The following case study illustrates how tidy data and tidy tools make data analysis easier by easing the transitions between manipulation, visualisation and modelling: you will not see any code that exists solely to get the output of one function into the right format for input to another. I'll show the \proglang{R} code that performs the analysis, but even if you're not familiar with \proglang{R} or the exact idioms I use, I've tried to make it easy to understand by tightly interleaving code, results and explanation.

@@

To achieve our goal of finding unusual time courses, we'll start by counting the number of deaths in each hour (\code{hod}) for each cause (\code{cod}) with the tidy \code{count} function. Then we remove missing (and hence uninformative for our purpose) values with \code{subset}. This gives Table~\subref*{tbl:counts:1}.

First, we count the the number of deaths in each hour of the day for each cause of death, and remove missing observations:

\begin{Verbatim}
hod2 <- count(deaths, c("hod", "cod"))
hod2 <- subset(hod2, !is.na(hod))
\end{Verbatim}

Next we join the dataset to \code{codes} dataset that gives informative labels for the disease. These two datasets are connected by the \code{cod} variable. This adds a new variable, \code{disease}, shown in Table~\subref*{tbl:counts:2}).

\begin{Verbatim}
hod2 <- join(hod2, codes, by = "cod")
\end{Verbatim}

The total deaths for each cause vary over several orders of magnitude: there were 46,794 deaths from heart attack, and 10 deaths from avalanche. This means that it makes more sense to compare the proportion of deaths in each hour, rather than the total number. We compute this by breaking the dataset down by \code{cod}, and then \code{transform}ing to add a new \code{prop} column, the hourly frequency divided by the total number of deaths from that cause. This new columns is Table~\subref*{tbl:counts:3}.

\code{ddply} breaks down its first argument (\code{hod2}) by its second argument (the \code{cod} variable), and applies its third argument (\code{transform}) to each piece. The fourth argument (\code{prop = freq / sum(freq)}) is passed on to transform.

\begin{Verbatim}
hod2 <- ddply(hod2, "cod", transform, prop = freq / sum(freq))
\end{Verbatim}

We then compute the overall average death rate for each hour, and merge that back into the original dataset. This yields Table~\subref*{tbl:counts:4} and allows us to easily compare each disease with the overall pattern by comparing \code{prop} to \code{prop\_all}.

First, we work out the number of people dying each hour by breaking down \code{hod2} by \code{hod}, summarising with a total for over each cause of death.

\begin{Verbatim}
overall <- ddply(hod2, "hod", summarise, freq_all = sum(freq))
\end{Verbatim}

Next we work out overall proportion of people dying in each hour:

\begin{Verbatim}
overall <- transform(overall, prop_all = freq_all / sum(freq_all))
\end{Verbatim}

Then finally join the overall dataset with the individual dataset to make it easier to compare the two:

\begin{Verbatim}
hod2 <- join(hod2, overall, by = "hod")
\end{Verbatim}

@@

  \caption{A sample of four diseases and four hours from \code{hod2} data frame.}
  \label{tbl:counts}
\end{table}

Next we compute a distance between the time course of each cause of death and the overall time course. Here we use a simple mean squared deviation.  We also record the sample size, the total number of deaths from that cause. To ensure that the disease we work with have a decent amount of underlying data we'll only 
work with diseases with more than 50 deaths in total ($\sim$2/hour).

\begin{Verbatim}
devi <- ddply(hod2, "cod", summarise, 
  n = sum(freq), 
  dist = mean((prop - prop_all)^2))

devi <- subset(devi, n > 50)
\end{Verbatim}

We don't know the variance characteristics of this estimator, so we explore it visually by plotting n vs.\ deviation, Figure~\ref{fig:deviation-raw}. The linear scale shows little, except the variability decreases with sample size, but on the log-log scale, Figure~\ref{fig:deviation-log} there is a clear pattern. It's particularly easy to see the pattern when we add the line of best fit from a robust linear model.

\begin{Verbatim}
ggplot(data = devi, aes(x = n, y = dist) + geom_point()

last_plot() + 
  scale_x_log10() + 
  scale_y_log10() +
  geom_smooth(method = "rlm", se = F)
\end{Verbatim}

@@

We are interested in points that are high relative to other causes with the same number of deaths: these are diseases most different from the overall pattern. To find these unusual points, we fit a robust linear model and plot the residuals, as shown in Figure~\ref{fig:devi-resid}. The plot shows an empty region around a residual of 1.5, so somewhat arbitrarily we select those diseases with a residual greater than 1.5. We do that in two steps: first selecting the appropriate rows from \code{devi} (one row per disease), and then finding the matching time course information in the original summary dataset (24 rows per disease).

\begin{Verbatim}
devi$resid <- resid(rlm(log(dist) ~ log(n), data = devi))
unusual <- subset(devi, resid > 1.5)
hod_unusual <- match_df(hod2, unusual)
\end{Verbatim}

Finally, we plot the time courses of the unusual causes in Figure~\ref{fig:disease}. We break the diseases into two plots because of the differences in variability: the top plot shows diseases with over 350 deaths and the bottom with less than 350. The causes of death fall in to three main groups: murders, drowning, and transportation related. Murders are more common at night, drowning in the afternoon, and transportation related during commute times. The pale gray line in the background shows the time course across all diseases.

\begin{Verbatim}
ggplot(data = subset(hod_unusual, n > 350), aes(x = hod, y = prop)) + 
  geom_line(aes(y = prop_all), data = overall, colour = "grey50") +
  geom_line() + 
  facet_wrap(~ disease, ncol = 3)
\end{Verbatim}
